---
title: "Analysis of the Impact of Severe Weather Events on Public Health and Economy"
author: "Andre Andrade Costa"
output: html_document
---

# Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

# Data Processing
The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. This data is available for download here:

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)


### Required Librarys

For this study the following  R packages are used:

```{r librarys, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(stringdist)
library(reshape2)
```



### Reading the data



```{r read, cache=TRUE}
if (!dir.exists("data")) {
    dir.create("data")
    download.file(
        url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
        destfile = "data/rawdata.csv.bz2")
}
data <- read.csv(bzfile("data/rawdata.csv.bz2"))
```


Size of data:
```{r, cache=TRUE}
dim(data)
```


Distinct event types in data set:
```{r, cache=TRUE}
length(unique(data$EVTYPE))
```

Top 50 events:
```{r, cache=TRUE}
byEvents <- data %>%
    group_by(EVTYPE) %>%
    summarise(count = n()) %>%
    arrange(desc(count))
head(as.data.frame(byEvents), 50)
```


Official events:
```{r, cache=TRUE}
events <- toupper(as.character(read.csv("event_types.txt", header = F)[,]))
events <- c(events, "OTHER")
print(events)
```


Classify events:
```{r, cache=TRUE}

data <- data %>% 
    mutate(EVTYPE = gsub("TSTM", "THUNDERSTORM", toupper(EVTYPE)),
           EVENTCLASS = events[amatch(EVTYPE, events, maxDist = 20, nomatch = 49)])

```


Events after classification:

```{r, cache=TRUE}
byEvents <- data %>%
    group_by(EVENTCLASS) %>%
    summarise(count = n()) %>%
    arrange(desc(count))
print.data.frame(byEvents)
```

Calculate damages
```{r, cache=TRUE}

data$PROPDMG_VAL = 0
data[toupper(data$PROPDMGEXP) == "H",]$PROPDMG_VAL = data[toupper(data$PROPDMGEXP) == "H",]$PROPDMG * 100
data[toupper(data$PROPDMGEXP) == "K",]$PROPDMG_VAL = data[toupper(data$PROPDMGEXP) == "K",]$PROPDMG * 1000
data[toupper(data$PROPDMGEXP) == "M",]$PROPDMG_VAL = data[toupper(data$PROPDMGEXP) == "M",]$PROPDMG * 1000000
data[toupper(data$PROPDMGEXP) == "B",]$PROPDMG_VAL = data[toupper(data$PROPDMGEXP) == "B",]$PROPDMG * 1000000000

data$CROPDMG_VAL = 0
data[toupper(data$CROPDMGEXP) == "H",]$CROPDMG_VAL = data[toupper(data$CROPDMGEXP) == "H",]$CROPDMG * 100
data[toupper(data$CROPDMGEXP) == "K",]$CROPDMG_VAL = data[toupper(data$CROPDMGEXP) == "K",]$CROPDMG * 1000
data[toupper(data$CROPDMGEXP) == "M",]$CROPDMG_VAL = data[toupper(data$CROPDMGEXP) == "M",]$CROPDMG * 1000000
data[toupper(data$CROPDMGEXP) == "B",]$CROPDMG_VAL = data[toupper(data$CROPDMGEXP) == "B",]$CROPDMG * 1000000000

```

# Results




You can also embed plots, for example:

```{r, cache=TRUE}

byHealth <- data %>%
    group_by(EVENTCLASS) %>%
    summarise(Total_Injures = sum(INJURIES), Total_Fatalities = sum(FATALITIES)) %>%
    mutate(Total_Health = Total_Injures + Total_Fatalities) %>%
    arrange(desc(Total_Health)) %>%
    top_n(10, Total_Health)
    

print.data.frame(byHealth)
    
byHealth_long <- melt(byHealth, id.vars=c("EVENTCLASS"), measure.vars=c("Total_Injures", "Total_Fatalities"))

```


```{r}

ggplot(byHealth_long, aes(x=reorder(EVENTCLASS, -value), y = value, fill=variable)) +
    geom_bar(stat="identity") + 
    labs(y = "Count of events", x = "Events", title = "Health Impact") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r, cache=TRUE}

byDamage <- data %>%
    group_by(EVENTCLASS) %>%
    summarise(Total_Property_Damage = sum(PROPDMG_VAL), Total_Crop_Damage = sum(CROPDMG_VAL)) %>%
    mutate(Total_Damage = Total_Property_Damage + Total_Crop_Damage) %>%
    arrange(desc(Total_Damage)) %>%
    top_n(10, Total_Damage)
    

print.data.frame(byDamage)
    
byDamage_long <- melt(byDamage, id.vars=c("EVENTCLASS"), measure.vars=c("Total_Property_Damage", "Total_Crop_Damage"))

```


```{r}

ggplot(byDamage_long, aes(x=reorder(EVENTCLASS, -value), y = value, fill=variable)) +
    geom_bar(stat="identity") + 
    labs(y = "US$", x = "Events", title = "Economy Impact") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
