---
title: "Analysis of the Impact of Severe Weather Events on Public Health and Economy"
author: "Andre Andrade Costa"
output: html_document
---

# Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

# Data Processing
The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. This data is available for download here:

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)


### Required Librarys

For this study the following  R packages are used:

```{r librarys, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(stringdist)
library(reshape2)
```



### Reading the data

In this section, the compacted file is downloaded and extracted and its content is read into a data frame.

```{r read, cache=TRUE}
if (!dir.exists("data")) {
    dir.create("data")
    download.file(
        url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
        destfile = "data/rawdata.csv.bz2")
}
data <- read.csv(bzfile("data/rawdata.csv.bz2"))
```


* Size of the data:
```{r, cache=TRUE}
dim(data)
```

The data is almost a hundred million records with 37 variables. But in this study we are interested only on the kind of the event, how many causalities (injuries and fatalities) and how much as the damage (property and crop). 

One major problem with this data set is that the events are not properly categorized.

* Distinct event types in the data set:
```{r, cache=TRUE}
length(unique(data$EVTYPE))
```


* Top 50 events:
```{r, cache=TRUE}
byEvents <- data %>%
    group_by(EVTYPE) %>%
    summarise(count = n()) %>%
    arrange(desc(count))
head(as.data.frame(byEvents), 50)
```

In the data documentation, is stated that there is only 48 known events. Here we manually list all known events and included the "OTHER" kind of event for the cases where we could not classify the event.

* Official events:
```{r, cache=TRUE}
events <- c("ASTRONOMICAL LOW TIDE", "AVALANCHE", "BLIZZARD", "COASTAL FLOOD", "COLD/WIND CHILL", "DEBRIS FLOW", "DENSE FOG", "DENSE SMOKE", "DROUGHT", "DUST DEVIL", "DUST STORM", "EXCESSIVE HEAT", "EXTREME COLD/WIND CHILL", "FLASH FLOOD", "FLOOD", "FREEZING FOG", "FROST/FREEZE", "FUNNEL CLOUD", "HAIL", "HEAT", "HEAVY RAIN", "HEAVY SNOW", "HIGH SURF", "HIGH WIND", "HURRICANE/TYPHOON", "ICE STORM", "LAKESHORE FLOOD", "LAKE-EFFECT SNOW", "LIGHTNING", "MARINE HAIL", "MARINE HIGH WIND", "MARINE STRONG WIND", "MARINE THUNDERSTORM WIND", "RIP CURRENT", "SEICHE", "SLEET", "STORM TIDE", "STRONG WIND", "THUNDERSTORM WIND", "TORNADO", "TROPICAL DEPRESSION", "TROPICAL STORM", "TSUNAMI", "VOLCANIC ASH", "WATERSPOUT", "WILDFIRE", "WINTER STORM", "WINTER WEATHER", "OTHER")
```


To clean up the events, first we substitute the acronym "TSTM" for "THUNDERSTORM" because in some events this acronym is used. Then we use the "stringdist" package so we could classify the events based on the similarity with the official events. Example: THUNDERSTORM WINDS -> THUNDERSTORM WIND:
```{r, cache=TRUE}

data <- data %>% 
    mutate(EVTYPE = gsub("TSTM", "THUNDERSTORM", toupper(EVTYPE)),
           EVENTCLASS = events[amatch(EVTYPE, events, maxDist = 20, nomatch = 49)])

```


* Events after classification:
```{r, cache=TRUE}
byEvents <- data %>%
    group_by(EVENTCLASS) %>%
    summarise(count = n()) %>%
    arrange(desc(count))
print.data.frame(byEvents)
```

According to the docs, the damage is recorded with the variables \___DMG and \___DMGEXP. So we created another variable that is the multiplication of the value with its order of magnitude:
```{r, cache=TRUE}

data$PROPDMG_VAL = 0
data[toupper(data$PROPDMGEXP) == "H",]$PROPDMG_VAL = data[toupper(data$PROPDMGEXP) == "H",]$PROPDMG * 100
data[toupper(data$PROPDMGEXP) == "K",]$PROPDMG_VAL = data[toupper(data$PROPDMGEXP) == "K",]$PROPDMG * 1000
data[toupper(data$PROPDMGEXP) == "M",]$PROPDMG_VAL = data[toupper(data$PROPDMGEXP) == "M",]$PROPDMG * 1000000
data[toupper(data$PROPDMGEXP) == "B",]$PROPDMG_VAL = data[toupper(data$PROPDMGEXP) == "B",]$PROPDMG * 1000000000

data$CROPDMG_VAL = 0
data[toupper(data$CROPDMGEXP) == "H",]$CROPDMG_VAL = data[toupper(data$CROPDMGEXP) == "H",]$CROPDMG * 100
data[toupper(data$CROPDMGEXP) == "K",]$CROPDMG_VAL = data[toupper(data$CROPDMGEXP) == "K",]$CROPDMG * 1000
data[toupper(data$CROPDMGEXP) == "M",]$CROPDMG_VAL = data[toupper(data$CROPDMGEXP) == "M",]$CROPDMG * 1000000
data[toupper(data$CROPDMGEXP) == "B",]$CROPDMG_VAL = data[toupper(data$CROPDMGEXP) == "B",]$CROPDMG * 1000000000

```


# Results

In this section we will analyse the health and economy impact of the just top 10 events.

First we select the top 10 events based on injuries and fatalities.
```{r, cache=TRUE}

byHealth <- data %>%
    group_by(EVENTCLASS) %>%
    summarise(Total_Injures = sum(INJURIES), Total_Fatalities = sum(FATALITIES)) %>%
    mutate(Total_Health = Total_Injures + Total_Fatalities) %>%
    arrange(desc(Total_Health)) %>%
    top_n(10, Total_Health)
    

print.data.frame(byHealth)

```

As we can see on the plot, tornadoes are by far the most dangerous kind of event. 

```{r}
byHealth_long <- melt(byHealth, id.vars=c("EVENTCLASS"), measure.vars=c("Total_Injures", "Total_Fatalities"))

ggplot(byHealth_long, aes(x=reorder(EVENTCLASS, -value), y = value, fill=variable)) +
    geom_bar(stat="identity") + 
    labs(y = "Count of events", x = "Events", title = "Health Impact") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Now we select the top 10 events based on property and crop damages.

```{r, cache=TRUE}

byDamage <- data %>%
    group_by(EVENTCLASS) %>%
    summarise(Total_Property_Damage = sum(PROPDMG_VAL), Total_Crop_Damage = sum(CROPDMG_VAL)) %>%
    mutate(Total_Damage = Total_Property_Damage + Total_Crop_Damage) %>%
    arrange(desc(Total_Damage)) %>%
    top_n(10, Total_Damage)
    

print.data.frame(byDamage)
    
```

Based on the plot, floods has the biggest impact on economy damages. 

```{r}

byDamage_long <- melt(byDamage, id.vars=c("EVENTCLASS"), measure.vars=c("Total_Property_Damage", "Total_Crop_Damage"))

ggplot(byDamage_long, aes(x=reorder(EVENTCLASS, -value), y = value, fill=variable)) +
    geom_bar(stat="identity") + 
    labs(y = "US$", x = "Events", title = "Economy Impact") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
